
## 1 - R Basics
\

Welcome to your first R tutorial! In this tutorial we'll be covering the very basics of R, which is a programming language originally developed for 
understanding data and doing statistics. This tutorial includes:  

* `R` packages
* the anatomy ("vocabulary" and "grammar") of R code
* objects
   + assignment to objects
   + object classes
* functions
   + function arguments
* outputs
* code comments
* loading data into `R`  
\
* What programming languages are
  + types of programming language
  + the anatomy ("vocabulary" and "grammar") of R code
* What RStudio is, and how RStudio works with R
* The structure of R
  + base R packages
  + add-on R packages
* Reading data into `R`.  
\  

With this information, we'll put our newfound knowledge into practice and 
do some basic coding in R to get used to this new way of working with data. If 
you're familiar with R already, this should still serve as a 
useful reminder!  
\ 

CONTINUE TO RESOURCES LIST...  


$~$

***
### RESOURCES
\

##### BOOKS
 
* Beckerman, Childs & Petchey (2017) _Getting started with R, an 
introduction for Biologists_, 2nd ed. pp 1 - 34, Chapter 1, "Getting and 
getting acquainted with R".  
\ 

##### CHEATSHEETS:

* [RStudio IDE cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/rstudio-ide.pdf) by RStudio
* [base-R cheatsheet](http://github.com/rstudio/cheatsheets/raw/master/base-r.pdf) by RStudio
\ 

##### WEB LINKS:

* [R for beginners](https://education.rstudio.com/learn/beginner) by RStudio  
\ 

Let's get started!  


$~$

***
### R PACKAGES
\

Lots of things can be done with what is called _base R_, 
which is a set of R functions and tools that load every time you start R 
(and therefore every time you start RStudio, as R is run from within RStudio).  

Some tasks require add-on _R packages_ with 
extra functions and tools. Think of these like dictionaries of extra vocabulary 
and grammar which allow R to communicate with the computer in more elaborate ways.  
\ 

This video explains briefly how `R` packages work:

<center>
![](https://youtu.be/-BLSElo-ZP4){width="80%"}
</center>
\  


$~$


***
### THE ANATOMY, VOCABULARY AND GRAMMAR OF `R` CODE
\

There are 6 key components to `R`code: objects, assignment, functions, 
arguments, outputs and comments.

<center>
![](https://youtu.be/-BLSElo-ZP4){width="80%"}
</center>
\  

##### **OBJECTS**
R is used for _Object-Oriented Programming (OOP)_. This means that when coding 
in R, you create **objects**, which are packets of data saved in the 
computer's RAM. These objects can be recalled and used throughout that R 
session, until they are either overwritten or removed.  

Think of it like filling a weekend bag full of things: your bag is full of 
objects ready to use; your `R` session is also full of objects ready for you 
to use.  

##### **CLASSES**
These objects have different **classes**, which define
what the object is and how `R` works with it. `R` will work with objects 
differently, depending on their class: for example, it is possible to calculate 
the mean of some numbers, but it's not possible to calculate a mean of some words!  

Similarly, you might have two similar-looking objects in your weekend bag, but 
their "class" would mean you use them in unique ways. Your toothbrush and your 
hairbrush are both brushes, but you (presumably) use them differently.

##### FUNCTIONS AND ASSIGNING
Objects are usually created using **functions**. These create or process data 
in some way and save it to the RAM (like putting something in the bag). The 
way this is done is using an **assigning arrow** `<-`. The object goes on the 
left, and the function on the right.  
`answer <- 6 * 7`



$~$

***
### OBJECTS
\

This section covers how to create objects, what classes mean, what functions 
are and how they work.  
\

Objects are created using **functions**. Simple functions include `+ - * /`, 
only one of which is not an Ed Sheeran album.

_In the exercise below, use these functions to:_

1. _Choose any number_
2. _Add the next highest integer number_
3. _Add 9_
4. _Divide by 2_
4. _Subtract your original number_

_Hint: don't forget the order of operations: which of `+ - * /` is done first!_
_Parentheses `( )` can help group the right operations._
```{r C01_simplefunctionsE1, exercise=TRUE, echo=FALSE}

```
```{r C01_simplefunctionsE1-solution}
# If your number is 42
((42 + 43 + 9) / 2) - 42

# Whatever number you chose, your answer is 5.
```
\  

More complicated functions take the form 
`function(argument1, argument2, ...)`. Each function has its own specific
set of arguments, and a different object is created depending on what 
these arguments are.
```{r}
numbers <- seq(1, 10, 1)
numbers
```
\

The function `seq()` means "sequence". This sequence is the integer numbers 
starting at 1 and ending at 10.  
\

We can do certain things with this vector. For example, we might want to see a 
summary about it, find out what its length is, calculate the mean or see the 
first 5 elements:
```{r}
summary(numbers)
length(numbers)
mean(numbers)
numbers[1:5]
```
\

In this case `1:5` does the same thing as `seq(1, 5, 1)`. There are often many 
ways to get any single output in `R`, in the same way that there are often 
many different ways to say one thing in a spoken language.  
\

We can change or add arguments to make a function behave differently. For example,
adding an argument _digits_ to the `summary()` function gives the results to 
a certain number of decimal places.
```{r}
summary(numbers, digits = 1)
```
\

Going back to the original creation of the `numbers` object, we can change the 
arguments of `seq` to create a different object: the two-times-table up to 20.
```{r}
numbers2 <- seq(2, 20, 2)
numbers2
```
\

_In the exercise below, create an object called `numbers3` which is a sequence_ 
_of numbers from 3 to 21, in steps of 2._
```{r C01_seqE1, exercise=TRUE, echo=FALSE}

```
```{r C01_seqE1-solution}
seq(3, 21, 2)
```
\  

As we have `numbers2` and `numbers3`, maybe we should be consistent and change 
the name of `numbers` to `numbers1`. We can do that by creating a new object and 
removing the old one:
```{r}
numbers1 <- numbers
rm(numbers)
```
\

We created `numbers1` from another object `numbers`. In fact, you can
create object using multiple other objects, and objects can be created in 
different ways.
```{r}
# We could create numbers2 by multiplying numbers1 by 2
numbers1 * 2

# We could add 1 to every number in numbers2
numbers2 + 1

# We could add together numbers1 and numbers2
numbers1 + numbers2
```
\

Note how `R` processes objects: both `numbers1` and `numbers2` have 10 individual 
numbers (termed _elements_ of the vector), 
so `R` adds the two first elements, the two second elements, the two third elements, 
etc. If we provide `R` with fewer than 10 elements, then it recycles the shorter 
vector: so if we add `1` to `numbers1` then it adds 1 to every element.  
\  



```{r C01_objectsQ1, echo = FALSE}
### *** TO COMPLETE!
quiz(caption = "Objects",
    question("What do the three arguments of the seq() function mean (in order)?",
        answer("start of sequence, length of sequence, sequence increment"),
        answer("start of sequence, end of sequence, length of sequence"),
        answer("start of sequence, end of sequence, sequence increment", correct = TRUE),
        incorrect = "start, end, increment.",
        random_answer_order = TRUE
    ),

    question("Here's a multiple choice question with more than one answer: which of these are amazing?",
        answer("Tea", correct = TRUE, message = "Tea is amazing."),
        answer("Coffee", correct = TRUE, message = "Coffee is amazing."),
        answer("Fruit tea"),
        answer("None of these"),
        incorrect = "Tea is amazing. Coffee is amazing. Fruit tea is cancelled.",
        random_answer_order = TRUE
    ),

    question("Here's a question which you can retry if you get the answer wrong. Are you excited to learn some R coding???",
        answer("No", message = "...wow. Harsh. Try again."),
        answer("Yes", correct = TRUE, message = "Yeah you are!!!"),
        allow_retry = TRUE,
        random_answer_order = TRUE
    )
)
```


$~$

***
### OBJECT CLASSES
\

Objects always have a **class** which defines how `R` processes and interacts with 
them. There are 5 different class types that it's important to know at this point.

* **numeric**
* **character**
* **factor**
* **logical**
* **data.frame**

### numeric
\

The objects `numbers1`, `numbers2` and `numbers3` have the class _numeric_: 
this means a set (called a _vector_) of 
numbers. You can see that the class of `numbers1` is numeric using the function 
`class()`:
```{r}
class(numbers1)
```
\

We have already seen a fair bit of how numeric vectors are processed by `R`.  
\


### character
\

We'll create a new object which looks kinda like `numbers1`, but is
not actually the same.
```{r}
characters1 <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10")
characters1
```
\

The `c()` function is really important. It means "concatenate", and basically 
glues together the arguments you give it. The `characters1` object 
looks similar to the `numbers1` object, but it's not. Note the quote marks 
around each vector element.  
\

_Exercise: find out the class of `characters1`._
```{r C01_characterE1, exercise=TRUE, echo=FALSE}
characters1 <- c("1", "2", "3", "4", "5", "6", "7", "8", "9", "10")

```
```{r C01_characterE1-solution}
# the answer is 'character'.
class(characters1)
```
\  

Let's try and find the same information about `characters1` as we did for `numbers1`.
```{r}
summary(characters1)
length(characters1)
mean(characters1)
characters1[1:5]
```
\

One of these functions went kinda wrong. As far as R is concerned, the 
elements of `characters1` are words, and it can't calculate the mean of some 
words. It returns `NA`, meaning "Not Applicable": R language for something which 
doesn't exist or is missing.  

The _Warning_ message makes sure you're aware that
the code did run and give an output, but something probably went wrong.  

R can also return an _Error_ message, which tells you that
the function has failed to run at all, and gave no output.  

_Pay attention to these messages_: usually if you get 
an Error or a Warning, much of your subsequent code will fail as it depends
on things being calculated properly.  
\

It's possible to convert things between different classes, although the 
behaviour is sometimes unpredictable so be careful!
```{r}
characters2 <- as.character(numbers2)
characters2
```
\


```{r C01_characterQ1, echo = FALSE}
# *** TO COMPLETE!
quiz(caption = "Objects",
    question("What do the three arguments of the seq() function mean (in order)?",
        answer("start of sequence, length of sequence, sequence increment"),
        answer("start of sequence, end of sequence, length of sequence"),
        answer("start of sequence, end of sequence, sequence increment", correct = TRUE),
        incorrect = "start, end, increment.",
        random_answer_order = TRUE
    ),

    question("Here's a multiple choice question with more than one answer: which of these are amazing?",
        answer("Tea", correct = TRUE, message = "Tea is amazing."),
        answer("Coffee", correct = TRUE, message = "Coffee is amazing."),
        answer("Fruit tea"),
        answer("None of these"),
        incorrect = "Tea is amazing. Coffee is amazing. Fruit tea is cancelled.",
        random_answer_order = TRUE
    ),

    question("Here's a question which you can retry if you get the answer wrong. Are you excited to learn some R coding???",
        answer("No", message = "...wow. Harsh. Try again."),
        answer("Yes", correct = TRUE, message = "Yeah you are!!!"),
        allow_retry = TRUE,
        random_answer_order = TRUE
    )
)
```
```{r C01_characterQ1-hint}
"Try running ?seq for the help page for the function (though R documentation is confusing at first!)"
```
\


### factor
\

A **factor** object is like a character object, but recognises groups of elements
which are the same, called _levels_. Let's take a look:
```{r}
factor1 <- rep(LETTERS[1:5], 2)
factor1 <- as.factor(factor1)
factor1
```
\

The object `LETTERS` is a character object which already exists in `R` without us having 
to create it. Note that it has to be LETTERS in capitals: `R` is case-sensitive, 
which means the code won't work if you have a lower-case letter rather than 
an upper-case letter or vice versa. This is a really common mistake to make and 
easy to miss!  

We use the square brackets to choose which letters we
want: we choose A, B, C, D and E, which are the first 5 letters of the alphabet.  

The `rep` function means repeat, so we repeat A to E five times, to give a vector 
that has a length of 25.  

Finally, we convert that character vector to a factor using `as.factor()` and 
overwrite the original `factor1` object by reassigning it.
```{r}
summary(factor1)
```
\

We can see from the summary that R recognises the five different factor levels,
A through E.  
\

```{r C01_factorQ1, echo = FALSE}
# *** TO COMPLETE!
quiz(caption = "Objects",
    question("What do the three arguments of the seq() function mean (in order)?",
        answer("start of sequence, length of sequence, sequence increment"),
        answer("start of sequence, end of sequence, length of sequence"),
        answer("start of sequence, end of sequence, sequence increment", correct = TRUE),
        incorrect = "start, end, increment.",
        random_answer_order = TRUE
    ),

    question("Here's a multiple choice question with more than one answer: which of these are amazing?",
        answer("Tea", correct = TRUE, message = "Tea is amazing."),
        answer("Coffee", correct = TRUE, message = "Coffee is amazing."),
        answer("Fruit tea"),
        answer("None of these"),
        incorrect = "Tea is amazing. Coffee is amazing. Fruit tea is cancelled.",
        random_answer_order = TRUE
    ),

    question("Here's a question which you can retry if you get the answer wrong. Are you excited to learn some R coding???",
        answer("No", message = "...wow. Harsh. Try again."),
        answer("Yes", correct = TRUE, message = "Yeah you are!!!"),
        allow_retry = TRUE,
        random_answer_order = TRUE
    )
)
```
```{r C01_factorQ1-hint}
"Try running ?seq for the help page for the function (though R documentation is confusing at first!)"
```
\


### logical
\

Logical objects are perhaps hard to understand at first, but they're simple
really. They basically indicate whether something is `TRUE` or `FALSE`. They
usually relate to a question we ask of our data.
```{r}
rand_norm <- rnorm(50, 0, 1)
over0 <- rand_norm > 0
over0
```
\

The result is information on which of your random data are greater than than 0.  
\

```{r C01_logicalQ1, echo = FALSE}
quiz(caption = "Random",
    question("If you ran the above code multiple times, would you get the same result each time?",
        answer("Yes", correct = TRUE),
        answer("No"),
        random_answer_order = TRUE
    ),

    question("You won't get the same result each time. Which part of the code causes that to happen?",
        answer("the function rnorm()", correct = TRUE, message = "The function rnorm() creates random numbers."),
        answer("The arguments 50, 0, 1", message = "The function rnorm() creates random numbers."),
        answer("The logical operator >", message = "The function rnorm() creates random numbers."),
        random_answer_order = TRUE
    )
)
```
\

There are lots of questions you can ask of your data using logical operators, and 
you will encounter many on your R journey e.g. 

* `>` greater than 
* `>=` greater than or equal to
* `<` less than
* `<=` less than or equal to  

These behave in exactly the way you would expect given how you've already seen 
`>` works.  

Logical operators (and `R` in general) don't always behave in the way you might 
expect, however. Take these logical operators:

* `==` exactly equal to
* `!=` not equal to

`==` tells you whether the two objects on the left and the right are exactly equal 
to each other:

```{r}
1 == 1
numbers2 == numbers1 * 2
numbers1 == characters1
```

We expect the first two to be true. But the final one also returns all `TRUE`
even though `numbers1` is numberic vector and characters1 is a character vector. 
All I can say is... sometimes `R` is weird.  

`!=` does the opposite job, telling us which elements on the right are NOT the same as 
the element(s) on the left.  
\

One final useful logical operator:

* `%in%` tells you which elements of one vector belong to another vector.
\

`%in%` is used to find out whether a value or values in one vector
(on the left) exist in another vector (on the right). For example:
```{r}
num2_in_num1 <- numbers2 %in% numbers1
num2_in_num1
numbers2[num2_in_num1]
```
\

If the logical object `num2_in_num1` is `TRUE` then that element of `numbers2`
is found in `numbers1`. When `num2_in_num1` is put in the square brackets it 
picks out all the elements of `numbers2` which are `TRUE` and shows them to us: 
these are the numbers in `numbers2` which are also in `numbers1`. It's not 
necessary when using `%in%` for the two vectors to have the same length.  
\

`==`, `!=` and `%in%` are VERY useful for handling data and subsetting data 
sets, and we'll see this later in the course.
\

_Exercise: put in the correct logical operators. The correct outputs are found in the_
_solutions._
```{r C01_logicalE1, exercise=TRUE, echo=FALSE}
# 1. ((42 + 43 + 9) / 2) - 42 is the same as ((31 + 32 + 9) / 2) - 31
((42 + 43 + 9) / 2) - 42     ((31 + 32 + 9) / 2) - 31

# the mean of one set of random numbers is NOT the same as the mean of another
mean(runif(100))     mean(runif(100))

# cake is greater than biscuits
"cake"     "biscuits"

# c("A", "B", "Z") are found in the last 10 letters of the alphabet
c("A", "B", "Z")     LETTERS[16:26]

```
```{r C01_logicalE1-solution}
1. TRUE, using ==
2. TRUE, using !=
3. FALSE, using >
This is another strange behaviour of R that I don't understand. I 
don't understand how or why one character object "cake" can be numerically greater than another
"biscuits", but R has some way of deciding. It works with any character object.
4. FALSE FALSE TRUE, using %in%
```
\  


### data.frame
\

Finally, we'll encounter the `data.frame`. This is the way most data is 
stored for data analysis. As we learned earlier, there's one column per variable
and one row per observation. `data.frame`s are simply collections of variables:
```{r}
data <- data.frame(numbers2, characters1, factor1)
data
```
\

To access variables in a data.frame, you simply use the `$` dollar sign.
```{r}
data$numbers1
```
\

Vectors are one-dimensional but data frames are two-dimensional. To quickly 
select a few rows and/or columns, we can use the square brackets but we need 
to provide two numbers: which rows first, followed by which columns.
```{r}
# If you want all columns then leave the second number blank.
# The first 5 rows of the data frame:
data[1:5, ]

# If you want all rows then leave the first number blank.
# The first and third columns of the data frame:
data[ , c(1, 3)]

# The first rows of columns 1 and 3:
data[1:5, c(1, 3)]
```
\

Having commas in the right places is important in `R`. If you don't have a 
comma to indicate both rows and columns, then you will encounter an error 
about "undefined rows" or "undefined columns". A missing or extra comma is a
really common mistake and easy to miss.  
\

You can read in `data.frame`s from your own data using the function 
`read.csv()`. Your data frame needs to be saved as a CSV file, in a data frame 
format. You can do this easily in Excel using File -> Save As -> CSV (Comma delimited).
Then you provide the path to the file. For example, on windows, if the 
file "data.csv" is stored in your "My Documents" folder, you use 
```{r, eval=FALSE}
read.csv("C:/Users/UserName/Documents/data.csv")
# (replace "UserName" with your user name)
```

On Mac, if the file was stored in your Documents folder then you may use 
```{r, eval=FALSE}
### *** CHECK
read.csv("~/Documents/data.csv")
```


$~$

***
### CREATING AN R SCRIPT
\ 

R code is saved in a file with the extension `.R`. In the same way as Micosoft 
Word knows to open files with `.docx` extensions and Adobe Reader knows to
open files with `.pdf` extensions, RStudio knows to open files with `.R` 
extensions, and to work with them in certain ways.  

Create an R file in the folder you wish to use as your working directory by 
following the steps in the video below.  
\ 

<center>
![](https://www.youtu.be/rWHV2VlQo2w){width="80%"}
</center>  
\ 

Save this file wherever you like.  

_This is the file in which you're going to write code to fulfil the tasks in Section 2 of each Chapter._ In each Chapter you will cover one aspect of the analytical process.
You'll apply the knowledge and skills learned during the Chapter to a biological data set and 
add to this code file, so that by the end of the 
book, you'll have completed a full analysis including Tidying and Visualising 
your data, fitting a model and evaluating how well it fits the data, 
tweaking the data, visualisations and the model if necessary, until you have some 
reliable and reproducible results.  


$~$

***
### FINAL TIPS
\

Use `?` to get help on a function. These help pages take a fair while to get used 
to, but they lay out what arguments a function takes, what those arguments 
are, what the output (value) of the function is, and some examples.  
\

We use the `#` (hash) symbol to create a comment: nothing after the `#` will be run
as code. It's good practise to comment your code as you go, so when you 
go back to it later, it makes sense to you. Trust me, you make life so much 
easier for your future self if you comment well!  
\

The internet is really useful. As you get used to R, you'll be able to find
blogs and articles and forums that give you answers. There are also R mailing
lists, but it's best not to email them until you're sure you've done everything
you can to solve your problem.  
\

If something goes wrong with your R coding (usually Errors), then 99.9% of the
time it's because you've made a mistake. R is very literal. Be careful!  
\

_Exercise: correct the mistakes ('bugs') in the lines of code below, bearing in_
_mind the common, small mistakes mentioned in this Chapter. Keep going until you get the_
_right outputs!_
```{r C01tipsE, exercise = TRUE, eval = FALSE}
# a sequence from 4 to 40 in steps of 4
mynums <- seq(4, 40, 4,)
mynums

# the first 5 and last 5 letters of the alphabet
mychars <- LETTERS[c(1:5, 21:26)]
# length of mychars
length(Mychars)

# turn mychars into a factor
myfacs <- asfactor(mychars)

# create data frame
mydata <- data.frame(mynums, mychars, myfacs)
# see numbers
mydata$mynum
# first 5 rows of data frame
mydata[1:5]
# first 2 columns of data frame
mydata[1:2, ]

```
\

These are really common mistakes to make. Be aware!


***
##### **END OF CHAPTER**
***





