
## 4 --- Data Management
\

_The first section of this Chapter will be completed at home before this week's online workshop._
This first section contains _Quizzes_ and _Exercises_. These multi-
choice questions and `R` coding exercises will test your learning on this chapter's
topic.

_The second section of this worksheet will be completed during the online workshop in your tutor groups._
This section contains some tasks for you to complete with a real data set, on the 
topic of this Chapter. You'll be coding this yourselves in `R` using RStudio.  
\

We've learned some basic stuff in R: objects, classes, assignments, functions,
arguments, outputs, comments. Importantly we've learned of a few common classes
of object that we'll often encounter during our R adventures: numeric, 
character, factor, logical, data.frame.  

We've also learned about __Tidy Data__. Tidy data allows you, as analyst padawans, 
to easily subset and manipulate data, so you're working with the data that you 
need to be working with.  

In this session, we're going to learn common means of manipulating data. 
We're going to acquaint ourselves with the 
[`dplyr`](https://dplyr.tidyverse.org/) package. `dplyr` is an expert 
data-wrangler which you can use to organise and manipulate data ready for 
plotting and analysis. We'll be learning:

* how to read data into R
* how to subset data
   + `slice()` and `filter()` to choose certain rows of the `data.frame`
   + `select()` to choose certain columns of the `data.frame`
* how to manipulate data
   + how to add a variable to the `data.frame` using `mutate()`
   + how to arrange the `data.frame` by variable values using `arrange()`
* how to use **pipes** `%>%` to make coding neater and cleaner.  
\

$~$

***
### **4.1 RESOURCES**  
\

##### **BOOKS**
 
<!-- ### *** Pages etc. -->
* [_**R for Data Science**_](https://r4ds.had.co.nz/) by Garrett Grolemund and Hadley Wickham. 
* [_**Getting Started with R: An Introduction for Biologists (2nd ed)**_](http://r4all.org/books/gswr2/) 
by Andrew Beckerman, Dylan Childs, and Owen Petchey, pp 35 - 78, Chapter 2, "Getting your 
data into R" and Chapter 3, "Data management, manipulation and exploration".
\ 

##### **CHEATSHEETS**

* data transformation with dplyr cheatsheet - 
see Resources section on blackboard or download 
[here](https://github.com/rstudio/cheatsheets/raw/master/data-transformation.pdf)
\

##### **WEB LINKS**

* [**dplyr package**](https://dplyr.tidyverse.org/)
* [**R for beginners**](https://education.rstudio.com/learn/beginner") by RStudio
* [**ShinyGLiM**](https://iainmstott.shinyapps.io/ShinyGLiM)  
\

##### **ShinyGLM**

In this Chapter, we'll mainly be working with the _DATA_ tab.  
\

***
### **4.2 R PACKAGES**  
\

In this Chapter we will 
need to use another R package developed by the RStudio team.

 * `dplyr`, is an `R` package which helps work with data: subsetting, transforming, manipulating... etc.

Remember that if you're working alongside this in RStudio, the package first needs to be installed onto your computer, and unless you update `R`, this only needs to be done _once_. 

```{r eval = FALSE}
install.packages("dplyr")
 ```
\

Then in order for `R` to know to use it in your session, it needs to also be loaded:
 ```{r eval = FALSE}
library(dplyr)
 ```

$~$

***
### **4.3 SUBSETTING DATA**
\  

On the DATA tab of the ShinyGLM app, you can see the games data.  

There are two panels on the right 
(although if your browser window is narrow or you're 
using a smartphone or a tablet then these might be at the bottom of the 
page): **DATA FRAME** and **CODE**. On the CODE tab, you'll see a couple of lines of code which, if you cast your mind back to _Chapter 1 R Basics_, will be familiar. These two lines of 
code are what you would need to use in R Studio to read in the games data and 
get R to show it.  

On the left (or at the top of the page if you're using a small screen or 
small browser window) are some options.  

These options allow you to subset the data:

* Subset rows either by...
  + row numbers using the `slice()` function, or
  + selecting variable values using the `filter()` function
* Subset columns either by...
  + column numbers using the `select()` function, or
  + variable names, also using the `select()` function.  
\

One HUGE advantage of `R` and other coding languages is how quickly and easily it is to manage your data in these ways. With a single line of code you can apply one or more criteria which strip out what you don't want and keep in what you do. When working with data, you usually find this shuffling and re-shuffling of the data is absolutely necessary.  

$~$

### 4.3.1 `slice()`
\

##### **CHOOSING ROWS BY NUMBER**

First we'll pick some rows by their number using `slice`. Imagine that we want to view only the data for group 2, which are rows 7-10 of the data.  

* Change the row numbers slider to values of **7** and **10**.
* Check that the DATA FRAME is displaying persons 7 to 10.
* Look at the CODE panel to see how it's changed.  
\

```{r qz431_3, echo = FALSE}
quiz(caption = "**< QUIZ 1 >** Subsetting Data - slice",
    question("How many objects are assigned in this code?",
        answer("1", correct = TRUE),
        answer("2"),
        answer("3"),
        answer("4"),
        message = "There are 3 objects: `data`, `rows` and `dataSubset`. Objects are assigned using the backwards arrow `<-`",
        random_answer_order = FALSE
    ),

    question("How could you specify the rows to subset without assigning the desired range to the object `rows`? Select one or more options.",
        answer("using `slice(seq(7, 10, 1))`", correct = TRUE),
        answer("using `slice(7:10)`", correct = TRUE),
        answer("using `slice(c(7, 8, 9, 10))`", correct = TRUE),
        message = "There's always more than one way of doing things in `R`. We've learned about all of these methods in previous chapters.",
        random_answer_order = TRUE
    ),

    question("Even though the above options require fewer lines of code, why might we want to use the `rows` object anyway?`",
        answer("`rows` can be re-used in later code", correct = TRUE),
        answer("We can change `rows` once, and the row values will change every time it's used in the code", correct = TRUE),
        answer("Using the `rows` object requires less computer RAM"),
        answer("Using the `rows` object makes you look cooler and better at coding"),
        random_answer_order = TRUE
    )
)
```
\

$~$

### 4.3.2 PIPE `%>%`
\

You'll notice that we use a **pipe** `%>%` in the code.

```{r eval = FALSE}
rows <- c(7, 10)

dataSubset <-games %>%
    slice( seq(rows[1], rows[2], 1) )
```
\

Whatever comes before the pipe is passed to whatever comes after the pipe. So, 
the chunk of code above could be interpreted as  

_"Take the `games` object (which is the games data frame), and use the `slice()` function to subset rows 7 to 10 from it._  

This is assigned to an object called `dataSubset`.  
\

There's another way we can write this code:

```{r eval = FALSE}
rows <- c(7, 10)

dataSubset <- slice(games, seq(rows[1], rows[2], 1) )
```
\

For something short like we're doing here, this might be OK. But as you 
need to do more and more things with your data, this gets more complicated:
you end up with functions inside functions, parentheses within parentheses.  
\

#### **< EXERCISE 1 >** Subsetting Data - pipe `%>%`
_Edit the code below so you condense the first three lines of code into one line. Hint: combine answers from "Subsetting Data: Quiz 1" and the information you just learned about pipes._
```{r ex432_1, exercise=TRUE, exercise.eval= TRUE, echo=FALSE}
rows <- c(7, 10)
dataSubset <- games %>%
    slice( seq(rows[1], rows[2], 1) )
dataSubset
```
```{r ex432_1-solution}
dataSubset <- slice(games, seq(7, 10, 1))
OR
dataSubset <- slice(games, 7:10)

Step 1: use slice(games, ...) instead of the pipe
Step 2: use seq(7, 10, 1) or 7:10 instead of rows object
Step 3: combine!
```
\

$~$

### 4.3.3 `select()`
\

The pipe `%>%` is particularly useful when you want to perform multiple operations
on an object: we'll see this now as we want to select not just rows, but columns 
from the data.

##### **CHOOSING COLUMNS BY NUMBER**

First, we'll choose by column number. Imagine that we don't want the data related to the online games: this is the columns `pballs`, `reaction` and `hardware` (columns 7, 8 and 9).
Back in the ShinyGLM app, with the rows 
still set to 7 and 10, pick _"Choose by... Column numbers"_ and 
change the slider to 1 and 6. Check the _DATA FRAME_ tab to check 
you have the right rows and columns, then check the _CODE_ tab. Notice that
there's a further step in the data subsetting, which uses `select()` to choose columns. The data should now show the lines for group 2 and the columns for everything except variables related to the online games.  
\

The pipes have become important here. The chunk of code above could be interpreted as:  

_"Take the `games` object [which is the games data frame], and use the `slice()` function to first, subset rows 7 to 10 and second, use the `select()` function to subset columns 1 to 6, from it."_  

Let's take a moment to imagine what that would look like without pipes:
\

```{r eval = FALSE}
rows <- c(7, 10)
cols <- c(1, 6)
dataSubset <- select(slice(games, seq(rows[1], rows[2], 1) ), seq(cols[1], cols[2], 1))
``` 
\

We also saw in _Chapter 1: R Basics_ that there is a way to use base `R` to 
select rows and columns from a data frame using square brackets `[]` (check back if you've forgotten). That 
would look like this:
\

```{r eval = FALSE}
rows <- c(7, 10)
cols <- c(1, 6)
dataSubset <- games[seq(rows[1], rows[2], 1), seq(cols[1], cols[2], 1)]
``` 
\

I hope you agree with me that the pipes make everything neater and easier to understand.  
\

#### **< EXERCISE 2 >** Subsetting Data - `select`
_Type in the code you get from the ShinyGLM app (from `# define row range` onwards). Try the following:_

1. _Swapping the `slice()` and `select()` lines_
2. _Deleting the white space at the start of the `slice()` and `select()` lines_
3. _Removing the new line after each pipe `%>%` so that the three lines are on the same line_
```{r ex433_2, exercise=TRUE, echo= FALSE}

```
\

```{r qz433_1, echo = FALSE}
quiz(caption = "**< QUIZ 2 >** Subsetting Data - `select()`",
    question("In this case, does it make a difference to the result whether the subsetting functions `slice()` or `select()` functions come first?",
        answer("Yes"),
        answer("No", correct = TRUE),
        message = "In this case, it doesn't make a difference whether rows or columns are selected first.",
        random_answer_order = FALSE
    ),

    question("Will it ever make a difference to the result which order subsetting functions are put?",
        answer("Yes", correct = TRUE),
        answer("No"),
        message = "If for some reason you use `slice()` or `select()` twice when subsetting your data, then the order those commands happen matters. If you only use `slice()` and `select()` once each, then the order of the subsetting commands don't matter.",
        random_answer_order = TRUE
    ),

    question("Does it matter whether you delete the white space at the start of the rows?",
        answer("Yes"),
        answer("No", correct = TRUE),
        message ="It takes some practice to know what parts of the code are essential and what parts are stylistic. In this case the white space helps draw quick attention to the fact that each line is 'passed' to the next one using the pipe... sort of like a series of steps. It's good practise in `R` coding.",
        random_answer_order = TRUE
    ),

    question("Does it matter whether there is a new line after each pipe or not?",
        answer("Yes"),
        answer("No", correct = TRUE),
        message = "Again, this is a stylistic choice, and is considered good practise.",
        random_answer_order = TRUE
    ),

    question("Do you think it would be possible to have the new line _before_ the pipe, so that ths `slice()` and `select()` lines start with a pipe?",
        answer("Yes"),
        answer("No", correct = TRUE),
        message = "Without a pipe at the end of the line, `R` considers the 'chain' of commands to be finished. This is an example of a coding practise which is fundamental, not stylistic.",
        random_answer_order = TRUE
    )
)
```
\

**It's important to note: pipes don't work with every function in `R`**. They work for these functions from the `dplyr` package that we encounter in this Chapter, but don't necessarily expect them to work otherwise.  
\

**It is bad practice to use row and column numbers to select data from a data frame.**

* If you add or remove rows or columns from your data file, then you have to change your code to match those changes.
* If you change the order of the rows or columns in your data file, then you have to change your code to match those changes.
\

##### **CHOOSING COLUMNS BY NAME**

`select()` can therefore also be used to subset according to column _**name**_. 
We're much less likely to make a mistake if we use those names rather 
than picking the wrong column number.  

In ShinyGLM, choose _COLUMNS | Choose by... Variable names_, and take out `pballs`, `reaction` and `hardware`. Note how the code then changes, 
and you're selecting variables by name instead of by number.  
\

#### **< EXERCISE 3 >** Subsetting Data - `select()`
_Edit the code below to give a data frame which has all columns except `group`. Hint: the result ShinyGLM gives you works fine, but is not the quickest way to do it; try googling something like "remove column dplyr select" and see if you can find a simpler solution._  

```{r ex433_3, exercise=TRUE, echo= FALSE}
dataSubset <- games %>%
    select()
dataSubset
```
```{r ex433_3-solution, echo= FALSE}
dataSubset <- games %>%
    select(-group)
dataSubset
```
\

$~$

### 4.3.4 `filter()`
\

Selecting rows using numbers is also not the best way to go about things. `filter()` is a function that can select rows according to variable values.  
\

##### **SELECTING ROWS USING VARIABLE VALUES**

In ShinyGLM, make sure that all columns are selected, and choose 
the option to select rows using variable values.  
\

**Logical** commands come into their own here. When we were selecting rows 7 through 10, we were effectively selecting only Group 2. If the number or order of the rows were to change, then rows 7 through 10 would no longer represent Group 2. Instead, we can use a logical command `group == G2` and the `filter()` function:

```{r}
dataSubset <- games %>%
    filter( group == 'G2' )
dataSubset
```
\
Try it yourself in ShinyGLM: with all columns selected, choose _ROWS | Choose by... Variable values_ and type `group == "G2"` in the box, and click REFRESH. You should have a data frame that includes only Group 2.  
\

Using `filter()` can be very useful for removing data outliers. There is an outlier in the `eyesight` variable, which is evident from the histogram for `eyesight`. We can remove this by excluding any values below this value.  
\

#### **< EXERCISE 4 >** Subsetting Data - `filter()`
_The `eyesight` variable has an outlier of 14. Edit the code below to remove the outlier value by using `filter()` to select values below this outlier, then plot a histogram for the `eyesight` variable using the new data frame (use the ShinyGLM app or look back on Chapter 2: Data Distributions if you're not sure)._
```{r ex434_4, exercise=TRUE, echo= FALSE}
# subset the data
dataSubset <- games %>%
    filter()
dataSubset

# plot the data

```
```{r ex434_4-solution, echo= FALSE}
# subset the data
dataSubset <- games %>%
    filter(eyesight < 14) # look at the data frame: the outlier is 14
dataSubset

# plot the data
hist <- ggplot(dataSubset, aes(x = eyesight)) +
    geom_histogram()
# This is the simplest histogram possible: you could choose to plot
# y as density, change the number of bins or bin width, fill colour,
# x axis limits, theme, etc., but you don't have to. The important 
# Thing is to use dataSubset as the data layer.

```
\

More questions on logicals below. If you're not sure of the answers then try them out in ShinyGLM. If you put in an invalid logical then the app will stop working and the screen will fade out. Don't worry: all you need to do is refresh it and try again.
\

```{r qz434_2, echo = FALSE}
quiz(caption = "**< QUIZ 3 >** Subsetting Data - `filter`",
    question("How would you choose everything _except_ group 2?",
        answer("`group != 'G2'`", correct = TRUE),
        answer("`group !%in% 'G2'`"),
        answer("`group =! 'G2'`"),
        answer("`group - 'G2'`"),
        message = "`!=` means 'not equal to'.",
        random_answer_order = TRUE
    ),
    question("How would you choose only short-sighted people (with negative values for `eyesight`?) Select all that apply.",
        answer("`eyesight < 0`", correct = TRUE),
        answer("`0 > eyesight`", correct = TRUE),
        answer("`eyesight <= 0`"),
        answer("`eyesight != 0`"),
        message = "logicals often work both ways. In this case `eyesight < 0` and `0 > eyesight` are equivalent.",
        random_answer_order = TRUE
    ),
    question("How would you choose only people that have been abroad in their lifetimes? Select all that apply.",
        answer("`countries >= 2`", correct = TRUE),
        answer("`countries > 1`", correct = TRUE),
        answer("`1 > countries`", correct = TRUE),
        answer("`countries <= 2`"),
        answer("`1 < Countries`", message = "careful of uppercase letters where they shouldn't be!"),
        message = "There are often several ways in which you can specify what you want using logicals!",
        random_answer_order = TRUE
    ),
    question("How would you choose only people who aren't adults?",
        answer("`role %in% c('child', 'baby')`", correct = TRUE),
        answer("`role = c('child', 'baby')`"),
        answer("`role == 'child', 'baby'`"),
        answer("`role == 'child' & 'baby'`"),
        random_answer_order = TRUE
    ),
    question("How would you choose every hardware except touchscreens?",
        answer("`hardware != 'touchscreen'`", correct = TRUE),
        answer("`hardware %in% c('mouse', 'keyboard', 'touchpad')`", correct = TRUE),
        answer("`hardware %not% 'touchscreen'`"),
        answer("`hardware == !'touchscreen'`"),
        random_answer_order = TRUE
    )
)
```
\

$~$

***
### **4.4 ADDING VARIABLES**

Sometimes we might want to add a variable to the data frame that's some kind of
function of one or more other variables.  

$~$

### 4.4.1 `mutate()`
\

`mutate()` is the function we'll use to add variables. If we wanted to know how many _extra_ continents someone had been to (i.e. how many continents they have been to, not including their home continent), then we need a variable which is continents minus 1.  

Select _ADD A VARIABLE..._ in ShinyGLM. This allows you to add a variable to the data. You need to give your new variable a name (e.g. `extra_continents`), and then tell `R` how to calculate it: `continents - 1`. If you click _ADD VARIABLE_ it'll add it to the data frame.  
\

#### **< EXERCISE 5 >** Adding Variables - `mutate()`
*Often when we analyse counts, we need to log-transform them. Add a variable called log_countries which is the natural log of `countries`. You'll have to look up how to calculate the natural log in `R`.*
```{r ex441_5, exercise=TRUE, echo= FALSE}
dataSubset <- games %>%
    mutate()
dataSubset
```
```{r ex411_5-solution, echo= FALSE}
dataSubset <- games %>%
    mutate(log_countries = log(countries))
dataSubset
```
\

You can use `mutate` as many times as you like to add as many variables as you like.  

$~$

***
### **4.5 CHAINING COMMANDS**
\

You can use the pipe to chain together as many commands as you like.  
\

#### **< EXERCISE 6 >** Chaining Commands
_Subset the data frame to include only data for students, add a variable with a name of your choice which represents `reaction` as a percentage rather than a proportion, remove all columns except `person`, `role`, `reaction` and your new percentage reaction variable._
```{r ex45_6, exercise=TRUE, echo= FALSE, exercise.cap="3.6 Chaining Commands: Quiz 1"}

```
```{r ex45_6-solution, echo= FALSE}
dataSubset <- games %>%
    filter(role == 'student') # could also use role %in% 'student'
        mutate(reaction_pc = reaction * 100)
            select(c('person', 'role', 'reaction', 'reaction_pc'))
# the order of the filter, mutate and select lines doesn't matter
dataSubset
```
\

$~$

***
### **4.X OTHER RESOURCES**
\

...

$~$

***
### **4.X FINAL TIPS**
\

...

$~$

***
### **< WORKSHOP >** Data Management
***  
\
...

$~$

***
##### END OF CHAPTER
***



<!--

***
### PART 2

We'll now put some of that learning into practise with a few tasks with the 
data. Try and complete these on your own, without consulting the app, and
I'm sure you copied the code over from the app as was suggested, so you can
edit that as you need to. First, pretty simple, but pretty important...  
\

_Task: read in the games data_  
\

You're interested in whether luck truly is a real thing in life, as you think
men are unluckier than any other gender. Drawing a picture card is unlucky, as 
picture cards look at you funny and you're pretty sure they're up to no good.
Drawing a black card is luckier than drawing a red card, as red is the colour 
of the devil. There are several steps to preparing your data...  

_Task: prepare your lucky, lucky data..._

* _You don't want to use any data from people who drew 4 cards, as 4 represents 
the 4 horsemen of the apocalypse. Remove any data where someone chose 4 cards._
* _You don't care about any variable that isn't associated with the cards, except
for the identity of the person: their gender and their group. Get rid of the 
variables you don't care about._
* _You need to know how many non-picture cards there were for each person, so
calculate that and add it to your data._
\

_Questions:_

* _How many rows of data do you have left?_
* _What function would you use to arrange the rows of the data frame by the number of 
picture cards drawn? (Hint: see the Getting Started with R book)... can you 
write the code that would do that?_
* _What's the mean total number of black cards drawn?_
* _What are the mean numbers of picture cards drawn, separately for each gender? 
Hint: you can look at p. 71 of Getting Started With R for this one, and we'll
encounter it later in the course as well._
\


***
### FINAL TIPS

That's a whistle-stop tour in data tidying. I'd encourage you to also look at 
p.75 of _Getting Started With R_ which shows how this information maps on to
the _base-R_ methods for working with `data.frame`s. Most people will use these, 
and they can sometimes be quicker and easier when you just have something 
very simple to do (this is a general rule with the `tidyverse` packages: 
often if you have something quick to do, then the base-R approach is better). 
Either approach is equally as valid.  
\

Note how I write R code: where there are spaces, what the indentation looks like,
etc. This is code _formatting_ and there's an accepted standard for what it 
should look like. Confusingly, it does matter in some instances where you do or 
do not put a space, but most of the time you can put a space (in fact as many spaces 
as you like), wherever you want. It's not the end of the world if you don't keep 
to the proper standards, but it does make code way easier to read.


-->