
<!-- spiel -->
As you complete Part 1 of this worksheet, you may find it useful to copy the code that the ShinyGLiM 
app shows, into the .R file you've been using for the 'games' data. Hopefully
you can create the same plots that the ShinyGLiM app shows you :)  

Remember again: typing code really is better than copy-pasting, as it helps the R language stick in your 
brain better.  

The first part of this worksheet contains _Questions_ and _Tasks_ in _italics_: note 
down the answers you get, and we'll go through them at the end of the session.  

The second part of this worksheet continues our adventures with our real data sets.
Having worked out how we need to subset the data, and looked at the distribution(s)
of the variable(s) we're working with, we're now going to see how those variables 
relate to one another. 
\

<!-- Resources -->
***
## RESOURCES

Books:
 
* Beckerman, Childs & Petchey (2017) _Getting started with R, an 
introduction for Biologists_, 2nd ed. pp 79 - 92, Chapter 4, "Visualizing
your data". Extra delectable content on pimping your plots on pp 203-218, 
Chapter 8, "Pimping your plots: scales and themes in `ggplot2`".
\ 

Cheatsheets:

* ggplot2 cheatsheet - see Resources section on blackboard or 
download [here](https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf)
\ 

Web links:

* [gpglot2](https://ggplot2.tidyverse.org/) by RStudio  
\ 

* [__ShinyGLiM__](https://iainmstott.shinyapps.io/ShinyGLiM)  
\


## PACKAGES

You'll need to load the right packages every time you complete worksheets. In 
this session, we will be working with:

 * `readr`, which helps load in data
 * `dplyr`, which is our data subsetting and manipulation friend
 * `ggplot2`, which makes pretty pretty data visualisations.
 ```{r echo = TRUE, message = FALSE}
library(readr)
library(dplyr)
library(ggplot2)
 ```
\


***
## PLOTTING MULTIPLE VARIABLES
Of course as scientists, what we really want to know is how a variable, let's 
call it the _dependent variable_, the _response variable_, the _y variable_ is
affected by another variable, which we could call the _independent variable_, the
_explanatory variable_, the _x variable_.  

In fact, those things are what we do call these variables, and there are more 
terms besides those. Different fields and different people like to use different 
terms. We're generally going to use the terms _dependent_ and _independent_, 
which generally refer to the _y_ and _x_ variables respectively. But
some points to consider are:

+ does the dependent variable depend on the independent variable? 
+ is the independent variable in fact independent?
+ does the response variable actually respond to the explanatory variable? If 
it doesn't, is the explanatory variable really explanatory?
+ do we plot the dependent variable on the y axis and the independent 
variable on the x axis?  

_Question: For each of the above questions, would you answer yes, or no?_  
\

In this session, we're going to reacquaint ourselves with the 
[`ggplot2`](https://ggplot2.tidyverse.org/) package. This time we'll be seeing
how variables relate to one another, and whether some variables may perhaps 
be influencing others. We'll be learning:

* another reinforcement of data-handling skills
* more about the grammar of graphics
* how plot two continuous variables against one another
   + scatterplots using `geom_point()`
* how to plot a categorical independent variable against a continuous dependent variable
   + box plots using `geom_boxplot()`
   + violin plots using `geom_violin()`
   + bar plots using `geom_bar()`  

We'll also learn a little more about customising plots.  
\


***
***
# PART 1

 \

## USER GUIDE

Reacquaint yourself with the USER GUIDE, as it may be able to answer some of 
your questions!  
\

## SUBSETTING DATA
For this exercise, our dependent variable is going to be the `ffold` variable.  

_Question: what statistical distribution does the `ffold` variable follow?_  
\

Our independent variables are going to be the `pballs` variable and the 
`ruler` variable (but we'll also see how other categorical variables impact 
a person's ability to forward fold). Our hypothesis is that people who have 
quick reactions and good aim must be sporty, and therefore be flexible.  

Because we're working with the `ffold` variable, we might want to cut out 
one group who didn't have that measured (group 5). That means we only want the first 
20 rows of data.  

_Question: pick the correct line of code to subset the data to the first 20 rows._
```{r eval=FALSE}
# a)
data %>% rows(c(1, 20))

# b)
data %>% slice(seq(1, 20, 1))

# c)
data %>% select(rows <= 20)

# d)
data %>% mutate(rows(1:20))

```
\

You can now use the app to check whether you were right by going to the 
DATA page, changing the row numbers slider and checking the CODE tab 
(the code may not be exactly the same as above, but hopefully you can see which 
function to use and that the two different lines of code will have the same result).  
\


## DATA VISUALISATION (TWO VARIABLES)
Now that we've subset the data, we can head right on to the DATA VISUALISATION 
(TWO VARIABLES) page. Choose `ffold` as your "dependent (y) variable". Note that:

* currently, the x variable is the first variable in the data, which should be 
`person`. 
* the "Plot type" is set to "points" (scatter plot).  

`person` is a categorical variable, but as you can see, it's still possible to 
plot a scatter plot anyway.

_Question: in this plot, what does each point represent?_  
\

We'd normally use something different like a box plot to plot a continuous 
dependent variable against a categorical x variable. Try selecting box plot from 
the "Plot type" list.  

_Question: the box plot should just have a single horizontal line for each 
person. Why is this?_  
\


### SCATTER (POINTS) PLOTS

We're interested, in fact, in the `pballs` and `ffold` variables. Choose 
`pballs` as the independent (x) variable. You should end up with a graph something 
like this:  
\

```{r echo = FALSE, warning = FALSE}
games <- read.csv("data/games.csv")
games_20 <- games %>% slice(1:20)
plot <- ggplot(games_20, aes(x = pballs, y = ffold)) +
    geom_point(color = 'steelblue4', fill = 'lightblue') +
    ylim(-15, 28) + ylab('ffold') + xlab('pballs') +
    theme_minimal()
plot
```

Thinking about our hypothesis that 
better aim will mean better flexibility...  

_Questions: what can we say about this graph, in terms of..._

* _what is the trend?_
* _does the trend support our hypothesis?_
* _are there any potential outliers or points with high influence over the trend?_  
\

OK, so this is a pretty beautiful graph, I think you'll agree.  

_Question: do you agree?_  
\

We can easily make this graph ourselves. Click on the CODE tab and you'll see a 
short chunk of code that makes this graph. If you're coding alongside this worksheet,
copy this over and see if you can recreate the plot.  

There are various bits to this code, and because code is quite literal, you can 
probably guess where to change things to make this plot look a little different. 
Here's a relatively big task: using your educated intuition...

_Task: copy this chunk of code and edit it so that:_

* _The y axis has limits of -20 and 30_
* _The points are "slategrey" colour_
* _The x axis has a label saying "Paper balls"_
* _The y axis has a label saying "Forward fold"_  
\

Now change those variables using the controls in the DATA VISUALISATION 
(TWO VARIABLES) page to see whether you got it right.  

Hint: you don't always have to include everything. We could exclude the 
`ylim()`, `ylab()`, `xlab()` and theme functions, and eliminate the arguments 
from the `geom_point()` function. We'd still get a graph, just plotted with all the
default arguments and values.
```{r echo = FALSE}
dataSubset <- games_20
```
```{r}
# initialise
plot <- ggplot(dataSubset, aes(x = pballs, y = ffold)) +
    # plot a histogram (geom)
    geom_point()

plot
```
\

_Question: What do you think this warning message could mean? (Hint: look 
at the data and pay attention to the `role` and any missing data in `ffold`)._  
\

We've made some conclusions about the link between a person's ability to 
aim paper balls at a wastepaper basket, and their ability to fold themselves in half.
Now we want to know whether someone's reaction time influences their ability 
to fold themselves in half. If you change the "independent (x) variable" to 
`ruler` then you should get a plot something like this:  
\
```{r echo = FALSE, warning = FALSE}
ggplot(games_20, aes(x = ruler, y = ffold)) +
    geom_point(color = 'maroon') +
    theme_minimal()
```
\

I changed the colour of the points, for a bit of a change. If you're coding 
alongside reading, you don't have to change the colour
(or please feel free to pick any colour you like!)  

Same _Questions_ here:

* _what is the trend?_
* _does the trend support our hypothesis?_
* _are there any potential outliers or points with high influence over the trend?_  
\

Although these are the variables we're interested in, there are other things that 
might affect the dependent variable, and they're worth checking out. Change the 
independent (x) variable to gender.  

_Question: Which grammar of graphics "layer" will this change in x variable affect?_

* _geometries_
* _data_
* _theme_
* _aesthetics_
* _facets_
\

Your graph should look something like this (but with a different colour):  
\
```{r echo = FALSE, warning = FALSE}
ggplot(games_20, aes(x = gender, y = ffold)) +
    geom_point(color = 'orange3') +
    theme_minimal()
```
\

Points are not really a helpful way to visualise data in groups, although 
we can sometimes tell something from them.  

_Question: which gender has the larger variance in `ffold`?_

* _female_
* _male_  
\


### BOX PLOTS

A more helpful way to look at these two variables is using a box plot:
```{r}
plot <- ggplot(dataSubset, aes(x = gender, y = ffold)) +
    geom_boxplot(color = 'goldenrod', fill = 'papayawhip') + 
        # some colour names are ridiculous...
    ylim(-15, 28) + ylab('Forward fold') + xlab('Gender') +
    theme_minimal()

plot
```
\

Our `fill` argument has actually come into some use now! (There are some 
points that have an outline and a fill: if you want to give it a try then try 
changing the `shape` argument in one of the earlier plots that use 
`geom_point()`). Some of the R colours have crazy names. You can also specify
colours using [hex values](https://www.w3schools.com/colors/colors_picker.asp)
if you like.  

The box plot shows us a little about the distribution of our data. If the 
quartiles are close to the median, then the distribution has a low variance.  

_Question: can a box plot also tell us something about the skewness? If so, how?_  
\


### VIOLIN PLOTS

Box plots are one of the oldest and most commonly used methods of visualising a 
continuous versus categorical variable, but `ggplot2` offers up another, newer 
one in the form of a "violin" plot. Choose "violin" as the plot type.  

_Question: take a guess at what the `geom` function will be to plot a violin
plot (i.e. points is `geom_point()`, boxplot is `geom_boxplot()`)..._

```{r echo = FALSE}
plot <- ggplot(dataSubset, aes(x = gender, y = ffold)) +
    geom_violin(color = 'darkgreen', fill = 'olivedrab2') + 
    theme_minimal()
```
\

Remembering back to the density plots of last week, a violin plot is essentially 
a density plot for different groups. Just as with other density plots, it has its 
drawbacks: you can see the whole distribution, but it's not possible 
to discern the moments of the distribution easily.  

It does seem that there are differences between the genders in their forward-folding
abilities. Perhaps not in the mean value, but in the variance at least. We may 
want to visualise three variables together: `pballs` or `ruler` as the independent 
variable, `ffold` as the depdendent variable, and then an extra grouping of 
`gender`. Think back to earlier plots with `pballs` or `ruler` on the x axis and 
`ffold` on the y axis.  

_Question: How could this plot be changed to also show the `gender` groups?_  
\

### BAR PLOTS

You may be familiar with another means of visualising categorical x continuous 
data, called a "barplot" (the fourth option). Please feel free to use bar plots to 
explore the data, but be aware that there are distinct disadvantages to
using bar plots, and many researchers are moving away from them nowadays. They 
describe very little in terms of the data variance or skewness, because they are just
bars showing the mean for each group. Additionally, because they require the y-axis to 
include 0, it can be difficult to see important effects in some data sets 
where values of the y variable are large, but variance, and differences between 
groups, are very small.  
\


### EXPERIMENTAL DESIGN, NONINDEPENDENCE AND BIAS

An important thing that we should start thinking about now is nonindependence and 
experimental design. In statistical analyses, each observation (data point) is usually 
assumed to be _**independent**_. This means that there is no way to predict any one 
data point from any of the other data points. __NOTE__: there is 
the concept of an independent _variable_ (the x variable), which is different 
to the concept of independent _data_ (data points not being dependent on one 
another). I know it's confusing, I'm sorry, I don't make the rules...  

_**Nonindependent**_ data is when we _can_ predict any one data point from 
another one. This is usually due to some way in which the data are collected, 
including spatial and temporal replication. In the 'games' data, there is 
nonindependence: each `group` has been tested at a different time, and/or 
working together and not with any of the other groups. `G1` is the guys; `G2` is the ladies, 
`G3` is some of your lecturers with whom I share an office, 
`G4` is my family, and further groups are incoming. Each group shares 
characteristics, may have taken a different approach to playing the games
(note that there are lots of things I _didn't_ tell you about how to play the games), 
have different ages or different variations in ages, and were tested at different
times of the day. All of these things could affect the data: if tested late 
in the day, reactions could be slower; if tested in the morning when muscles 
are stiffer, people could be worse at the forward 
fold than other groups tested in the evening; different groups may have 
thought of different ways to do each task... What this means is that data 
within each group may be likely to be more similar to one another than to 
other data in other groups.  

_Task: plot `ffold` as the dependent variable and `group` as the independent 
variable. Is there any evidence that data points within groups might 
actually be similar to one another in any way?_  
\

Think also about the "experimental design" in the way the data was collected. 
There may be biases if one data point was collected in a slightly different 
way to another, when they should actually have been collected in exactly the 
same way.

_Task: we'll have a discussion later about nonindependence and biases in the 
'games' data. Think and write down possible ideas about where nonindependence 
and biases appear in the data._  
\


***
***

# PART 2

We're back to working with your data! Use the same .R file as you used for 
single variable visualisation, and we will carry on from where we left off. 
If you've still got some work to do on last week's tasks, please go for it and let
me know if you need a hand.  

By the end of last week's tasks, you should have: 

* subset the data to remove any rows you don't need
* Added any extra variables (e.g. log-transformations), if you need them
* Plot the distributions of numeric variables to assess their bounds, 
* symmetry, skewness, and other characteristics (if any).
\

_Tasks:_

* _according to your hypothesis, identify which variable is your dependent variable._
* _according to your hypothesis, identify which variable is your key independent variable._
* _identify any further grouping (categorical) variables which may influence the dependent variable._
* _in R, using `ggplot2`, create the most appropriate plots to show:_
   + _what relationship (if any) there is between the dependent and independent variable_
   + _whether and how any other grouping (categorical) variables affect the dependent variable_

Pay attention to trends in the data, spread of the data, potential outliers, 
differences in the mean values and distributions within and between different groups, 
and come up with a preliminary statement on whether the plots seems to initially 
support the hypothesis, and whether any other variables may need to be taken into account.
\
\
