--- 
title: "BGY2011M Data Skills for the Life Sciences"
author: "Module lead: IAIN STOTT  /   email istott@lincoln.ac.uk"
date: "Last updated `r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
layout: default
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This workbook accompanies the University of Lincoln School of Life Sciences Level 2 Module BGY2011M Data Skills for the Life Sciences."
---

***
# Introduction 

Welcome to Data Skills for the Life Sciences! This page contains important 
information on the module: it's worth reading through carefully, and the 
information will always be here for you to refer back to. The first few sections
describe everything you need to know to set yourself up for the course. It's 
really important you complete these as soon as possible. Information on the 
timetable and learning schedule, and both formative and summative assessments, 
are found at the bottom of the page.


***
### The analytical process

This module is all about analysing data. 
Any analytical process is really just a means of understanding your data, and 
relationships between variables in that data. _That's why I do not consider this_
_to be a course in statistics; it is a course in giving you the skills to_
_understand data._ Statistics is one part of that.  

It's important not to dive right into statistical analysis without 
understanding your data, your variables and the relationships between them
first. For that reason, much of the course will be dedicated to understanding 
data, and how to plot data.  
These steps forms the **FIRST HALF of the course**:

* **Tidy data**
  - data exploration
  - data subsetting
  - data manipulation
* **Plots**
  - Single-variable distributions and summary statistics
  - Two- and three-variable plots to explore hypotheses

Data handling and plotting is also the best vehicle to learn how to 
code: data handling and plots are harder to code but easier to interpret, 
whereas statistical analyses are easier to code but harder to interpret.  

Equally, you cannot complete a statistical analysis without assessing how well 
that model fits the data, and whether any 'assumptions' of your statistical 
model are violated.  
These steps form the **SECOND HALF of the course**:

* **Model**
  - fitting a statistical model
* **Model fit**
  - assessing how well a statistical model fits the data
* **Results**
  - Understanding and interpreting the results of a statistical model

The "statistical model" we will be working with is the 
_general linear model (glm)_. This is a framework which brings together 
many traditional statistical analyses. You will learn where these traditional 
analyses fit into the glm framework.  

There is naturally feedbacks between these steps, where new information may 
mean you need to revise earlier steps.  
\ 

![The analytical process and its feedbacks, from start to finish][analytical_process]

[analytical_process]: Process.jpg "The Analytical Process"  
\ 


## Installing R and RStudio

Install here

```{r, eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<iframe style="position: relative; height: 100%; width: 100%;"
  src="https://www.youtube.com/embed/d7lARa168gk"
  frameborder="0" allow="accelerometer; autoplay; encrypted-media;
  gyroscope; picture-in-picture" allowfullscreen>
  </iframe>'
)
```



## Installing R packages

PACKAGES


## iain's notes

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
