--- 
title: "BGY2011M Data Skills for the Life Sciences"
author: "Module lead: IAIN STOTT  /   email istott@lincoln.ac.uk"
date: "Last updated `r Sys.Date()`"
site: bookdown::bookdown_site
output: bookdown::gitbook
documentclass: book
layout: default
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
github-repo: rstudio/bookdown-demo
description: "This workbook accompanies the University of Lincoln School of Life Sciences Level 2 Module BGY2011M Data Skills for the Life Sciences."
---


```{r echo = FALSE}
# packages
library(ggplot2)
```

# Introduction 

Welcome to Data Skills for the Life Sciences! This page contains important 
information on the module: it's worth reading through carefully, and the 
information will always be here for you to refer back to. The first few sections
describe everything you need to know to set yourself up for the course. It's 
really important you complete these as soon as possible.  
\ 

***
## This book

First, some **information on this web book**. This book accompanies every student for 
the first 9 weeks of the module, after which the teaching and content start
to diverge for different programmes. This allows programs to learn different 
analytical techniques which are relevant to their courses: some stay with this 
content, and some do not. The learning and assessment schedule can be found at 
the bottom of the page.  

### Using the web book

You will see some icons at the top of the page:  

![][gitbook_icons]

[gitbook_icons]: gitbook_icons.jpg "Icons"  

From left to right these are: 

* the _hamburger_ icon on the left shows or hides the page navigation
sidebar. 
* The _magnifying glass_ icon allows you to search the book for any 
term you like. 
* The _format_  (A) icon is designed for accessibility and allows you to: 
  - increase or decrease font size,
  - change the font from sans serif to serif, depending on what you find easiest
  to read,
  - change the colour scheme to black-on-white, toned-down sepia or white-on-
  black, depending on what you find easiest to read.
* The _edit_ icon in the middle is not of interest (it redirects any 
other web developers to a site where they can copy and edit the book). 
* The _download_
button allows you to download the book as a pdf (e.g. for printing out), or an epub 
(for reading on an eReader). Videos and interactive content will NOT work in pdf 
or epub formats (although these formats will contain relevant URLs). Therefore, 
this web version should ideally be the one you work from primarily, and it will work 
on any internet-enabled device with a screen through a web browser (although 
perhaps smartphone screens are a little small). 
* The _information_ icon gives 
some useful information on keyboard shortcuts.  

### Web book contents

Within the text you will find a diversity of content. You'll get very familiar
with R code! There's an example of some R code below. Code (of any type) 
is always written using a `fixed width font` like Courier New, Consolas or 
Lucida Console (my personal favourite at the moment is
[Space Mono](https://fonts.google.com/specimen/Space+Mono)). The reason for this 
is that all the letters line up vertically, which is far easier to read.  

_Code chunks_ are collections of lines of code which do a certain thing. There's 
an example of a code chunk below. It's just for you to see what code chunks and 
their outputs look like, so you don't need to understand it just yet! Here's 
what it does: 

* draw 1000 random numbers from a normal distribution with a mean of zero 
and standard deviation of 1
* create a simple plot of the distribution
* calculate the mean and standard deviation of the random distribution (which 
of course should be approximately zero and one respectively):  

```{r}
random <- rnorm(1000, 0, 1)
ggplot(as.data.frame(random), aes(x = random)) + 
    geom_density()
mean(random)
sd(random)
```
\ 
_Scripts_ are files which contain many chunks of code.  

Running R code results in many outputs, especially text outputs (such as th)

The worksheets also include instructional videos (a couple of examples are 
found further down on this page) text and plots, and interactive 
quizzes. Each chapter is arranged like this:

* **Part 1** of each chapter uses a web application I wrote to help teach analysis 
using R. It is available at 
[iainmstott.shinyapps.io/ShinyGLiM](iainmstott.shinyapps.io/ShinyGLiM). 
By interacting with the app, you'll learn how to code different tasks and 
write these into a script which ultimately will contain a full analysis of a
data set. 
* **Part 2** of each chapter is an interactive quiz which tests your knowledge of the 
topics and skills learned in Part 1. Here is an example:  

```{r, include=knitr::is_html_output(), eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<div style = "overflow: hidden; padding-top: 66.66%; position: relative;">
  <iframe style="border: 0; height: 100%; left: 0; position: absolute; top: 0; width: 100%;"
    src="https://shiny.rstudio.com/gallery/kmeans-example.html">
  </iframe>
</div>'
)
```

```{r, include=knitr::is_html_output(), eval=knitr::is_html_output(excludes = c("html", "html5")), results = 'asis', echo = F}
cat(
'Quiz example: https://shiny.rstudio.com/gallery/kmeans-example.html'
)
```

```{r, include=knitr::is_latex_output(), results = 'asis', echo = F}
cat(
'Quiz example: https://shiny.rstudio.com/gallery/kmeans-example.html'
)
```
\ 

* **Part 3** of each chapter presents some analytical tasks for biological data 
for you to code in R.  
\ 


### Learning schedule 

Parts 1 and 2 of each chapter are to be completed at home each week, prior to the 
online session. These should take 1-2 hours per chapter, although this will 
definitely vary depending on the topic. Part 3 is to be 
completed during the online sessions, when you will work in your tutor groups. 
You may complete Parts 1 and 2 either individually, or in tutor groups, as you wish.  

Information on the timetable, and both the formative and summative assessments, 
are found at the bottom of the page.  
\ 


***
## The analytical process

This module is all about analysing data. 
Any analytical process is really just a means of understanding your data, and 
relationships between variables in that data. _That's why I do not consider this_
_to be a course in statistics; it is a course in giving you the skills to_
_understand data._ Statistics is one part of that.  
\ 

![The analytical process and its feedbacks, from start to finish][analytical_process]

[analytical_process]: Process.png "The Analytical Process"  
\ 

There is a process to analysis, with a few different steps and feedbacks between 
them, as illustrated above. It's important not to dive right into statistical 
analysis without understanding your data, your variables and the relationships 
between them first. For that reason, much of the course will be dedicated to 
understanding data, and how to plot data.  

These steps forms the **FIRST HALF of the course**:

* **Tidy data**
  - data exploration
  - data subsetting
  - data manipulation
* **Plots**
  - Single-variable distributions and summary statistics
  - Two- and three-variable plots to explore hypotheses

Data handling and plotting is also the best vehicle to learn how to 
code: data handling and plots are harder to code but easier to interpret, 
whereas statistical analyses are easier to code but harder to interpret.  

Equally, you cannot complete a statistical analysis without assessing how well 
that model fits the data, and whether any 'assumptions' of your statistical 
model are violated.  

These steps form the **SECOND HALF of the course**:

* **Model**
  - fitting a statistical model
* **Model fit**
  - assessing how well a statistical model fits the data
* **Results**
  - Understanding and interpreting the results of a statistical model

The "statistical model" we will be working with is the 
_general linear model (glm)_. This is a framework which brings together 
many traditional statistical analyses. You will learn where these traditional 
analyses fit into the glm framework.  

There are naturally feedbacks between these steps, where new information may 
mean you need to revise earlier steps.  
\ 

***
## Installing R and RStudio

We will be using a program called _RStudio_ (the technical term for it is an 
"Integrated Development Environment" or _IDE_), to "code" in a programming 
language called _R_. The very first thing you need to do is install both of 
these, and the following video walks you through that:  


```{r, include=knitr::is_html_output(), eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<div style = "overflow: hidden; padding-top: 56.25%; position: relative;">
  <iframe style="border: 0; height: 100%; left: 0; position: absolute; top: 0; width: 100%;"
    src="https://www.youtube.com/embed/d7lARa168gk"
    frameborder="0" allow="accelerometer; autoplay; encrypted-media;
    gyroscope; picture-in-picture" allowfullscreen>
  </iframe>
</div>'
)
```

```{r, include=knitr::is_html_output(), eval=knitr::is_html_output(excludes = c("html", "html5")), results = 'asis', echo = F}
cat(
'Installing R and RStudio video: https://youtu.be/d7lARa168gk'
)
```

```{r, include=knitr::is_latex_output(), results = 'asis', echo = F}
cat(
'Installing R and RStudio video: https://youtu.be/d7lARa168gk'
)
```
\ 


***
## Why R?

You may have come across other programs that allow you to work with data, 
plot data and run statistical analyses. So why a programming language, and why 
R?  

There are a huge number of benefits to using R, listed here and elaborated on in 
the video below.

1. It's **_free_**.
2. It's **_open source_**.
3. It's **_cross-platform_**.
4. It's **_transparent_** and **_replicable_**.
5. It's **_powerful_** and **_fast_**.
6. It's **_flexible_**.
7. It makes you **_employable_**.  

Think of R as a language you're writing in, and 
RStudio like Microsoft Word, i.e. a program that allows you to write, 
to edit files of different types, to save them and generally to work with them.  

Those files don't do anything by themselves, but "running" R code from within a file allows you 
to speak to the computer and instruct it to do something. The computer will accordingly 
complete certain tasks like calculate something, create a plot or do a statistical 
analysis.  

RStudio enables you to write code, run code and view the outputs of 
your hard labour!  

Lots of things can be done with what is called _base R_, 
which is a set of R functions and tools that load every time you start R 
(and therefore every time you start RStudio, as R is run from within RStudio). 
However some tasks require add-on _packages_ with 
extra functions and tools: think of these like dictionaries of extra vocabulary and grammar 
which allow R to communicate with the computer in more elaborate ways.  

This video explains some of these concepts:  

```{r, include=knitr::is_html_output(), eval=knitr::is_html_output(excludes = "epub"), results = 'asis', echo = F}
cat(
'<div style = "overflow: hidden; padding-top: 56.25%; position: relative;">
  <iframe style="border: 0; height: 100%; left: 0; position: absolute; top: 0; width: 100%;"
    src="https://www.youtube.com/embed/-BLSElo-ZP4"
    frameborder="0" allow="accelerometer; autoplay; encrypted-media;
    gyroscope; picture-in-picture" allowfullscreen>
  </iframe>
</div>'
)
```

```{r, include=knitr::is_html_output(), eval=knitr::is_html_output(excludes = c("html", "html5")), results = 'asis', echo = F}
cat(
'Installing R and RStudio video: https://youtu.be/-BLSElo-ZP4'
)
```

```{r, include=knitr::is_latex_output(), results = 'asis', echo = F}
cat(
'Installing R and RStudio video: https://youtu.be/-BLSElo-ZP4'
)
```
\ 


```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
